---
title: "Text Mining - Exploration of a Text Corpus with R"
author: "Tomás E. Tecce"
date: "March 20, 2016"
output:
  html_document:
    theme: readable
    highlight: pygments
references:
- id: tm_r
  title: Text mining infrastructure in R
  author:
  - family: Feinerer, I.
  - family: Hornik, K.
  - family: Meyer, D.
  container-title: Journal of Statistical Software
  volume: 25
  URL: http://www.jstatsoft.org/v25/i05
  page: 1-54
  type: article-journal
  issued:
    year: 2008
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

In this report I describe how to load, pre-process and explore a dataset
of text documents using the `tm` package in the R programming language.
The final objective of this project is to use these documents as training
for a text predicting model. After removing frequent words, numbers and
punctuation, I calculate the frequency of single words and pairs of words
(two-grams) in the documents. 


## Source data

The data for this project comes from HC Corpora
(http://www.corpora.heliohost.org/index.html), a collection of corpora for
various languages freely available to download. The texts in the corpora
are collected from publicly available sources by a web crawler. 

The files for this project must be downloaded from the link provided in
the course website. After extracting the downloaded zip file, we end up
with four sets of three files:

- `aa_AA.blogs.txt`: text obtained from blogs
- `aa_AA.news.txt`: text obtained from news feeds
- `aa_AA.twitter.txt`: text obtained from Twitter

where `aa_AA` denotes language and locale and is either `de_DE` (German),
`en_US` (English, US), `fi_FI` (Finnish) or `ru_RU` (Russian). **For this
initial round of exploration I'll only use the English language files.**


## Selecting a data sample

The data files are relatively large (between 150 and 200 MB) containing
from about 900,000 lines of text in the blogs file to over 2 million in
the Twitter file (see the Appendix for detailed code and results). In
order to speed up data exploration and the development of an initial,
prototype model, I'll use smaller samples from each of the files. **In
this report, for illustration purposes, I'll use a sample of approximately
10,000 lines of text extracted at random from the news file** (1 per cent
of the total number of lines in the file). I choose the news file for this
initial exploration because it contains shorter lines than the blogs file
and also because I expect the text it contains to be more gramatically 
correct than texts in the other two files (especially the Twitter one).

The lines of text are extracted using the `rbinom` function of R, and to
ensure reproducibility the sample is saved to a file in plain text format 
(see the Appendix for the corresponding code.)


## Data preprocessing

For this initial exploration of the data I use the `tm` text mining
package [@tm_r]. First I load the data into a *Corpus* (a collection of
documents) which is the main data structure used by `tm`.

```{r create_corpus}
library(tm)

createCorpus <- function(filepath) {
  conn <- file(filepath, "r")
  fulltext <- readLines(conn)
  close(conn)
  
  vs <- VectorSource(fulltext)
  Corpus(vs, readerControl=list(readPlain, language="en", load=TRUE))
}

news_corpus <- createCorpus("data/samples/news_sample_report.txt")
```

The result is a structure of type `VCorpus` ('virtual corpus' that is,
loaded into memory) with 10,148 documents (each line of text in the source
is loaded as a document in the corpus). One thing I notice at this stage
is that the text file, when loaded into R, occupies 2.5 MB whereas the
associated VCorpus object is much larger, at 38.6 MB. This means that I
may have to find out another method when building the predictive model, as
converting the training datasets to VCorpus objects may not be possible
due to memory constraints.

Common data cleaning tasks associated with text mining are:

- Converting the entire document to lower case
- Removing punctuation marks (periods, commas, hyphens etc)
- Removing stopwords (extremely common words such as "and", "or", "not",
"in", "is" etc)
- Removing numbers
- Filtering out unwanted terms
- Removing extra whitespace

The `tm` package provides several function to carry out these tasks,
which are applied to the document collection as transformations via the
`tm_map()` function.

Keeping in mind that our final goal is to develop a text prediction
application, here I'll discuss whether the steps listed above are
appropriate in this case.

### Converting text to lower case

Converting to lower case is definitely useful since I won't want to
differentiate between e.g. "car", "Car" and "CAR". 

```{r to_lowercase}
news_corpus_proc <- tm_map(news_corpus, content_transformer(tolower))
```

### Removing stopwords

Removing stopwords is also very convenient in principle, although I'm not
too certain. Stopwords are words so common that their information value is
very low. To check how common they are in our sample, we do a simple word
count exercise for a small set of stopwords in the code chunk below.

```{r count_words, cache=TRUE}
countWords <- function(filepath, pattern) {
    conn <- file(filepath, "r")
    fulltext <- readLines(conn)
    close(conn)
    
    count <- 0
    for (i in 1:length(fulltext)) {
        findr <- gregexpr(pattern, fulltext[i])
        if (findr[[1]][1]>0) {
            count <- count + length(findr[[1]])
        }
    }
    count
}

news_file <- "data/samples/news_sample_report.txt"

# Adding the length of the file (in lines) to count the last word in each
# line, which is not surrounded by spaces
totwords <- countWords(news_file, " * ") + 10148

# Count the ocurrences of very common stopwords
mystopwords <- c(" [Aa]nd ", " [Ff]or ", " [Ii]n ", " [Ii]s ", " [Ii]t ",
                 " [Nn]ot ", " [Oo]n ", " [Tt]he ", " [Tt]o ")
totstops <- sum(sapply(mystopwords,
                       function(x) { countWords(news_file, x) }))
totstops/totwords
```

That is, **the small list of stopwords considered here accounts for almost
16 per cent of the total words in the sample data**. If we let them in the
sample, when considering two- and three-grams (sequences of 2 or 3
consecutive words) two-grams consisting of e.g. "and the", "for the", "or
the" will overwhelmingly dominate the rest of possible two-grams. I prefer
to start with a model that attempts to predict without the use of
stopwords, and will consider adding them to the training data if the
performance of the first model needs improvement.

Fortunately I do not need to compile a list of all possible stopwords - 
the `tm` package already includes a collection of stopwords for several
different languages.

```{r remove_stopwords}
news_corpus_proc <- tm_map(news_corpus_proc, removeWords,
                           stopwords(kind="en"))
```

### Removing punctuation and numbers

Removing punctuation marks may generate problems. For example, removing
periods from the data may result in $n$-grams formed with words from
different sentences. There is also the issue of the apostrophes, which in
English are used for abbreviations and possesives. Again, I'll ignore this
issue for the moment; I will later return to it as a possible way of
improving model performance.

In the case of numbers (e.g. "5", not the word "five") I think it would
be very difficult to predict the actual number the user would enter after
a phrase such as "I am going to buy" (if a number is entered at all).
Therefore, for the initial approach all numbers will be removed from the
training data.

```{r remove_punctuation}
news_corpus_proc <- tm_map(news_corpus_proc, removePunctuation)
news_corpus_proc <- tm_map(news_corpus_proc, removeNumbers)
```


### Removing unwanted terms

The source data contains plenty of profanity, but I don't want the final
application to suggest foul language to the users. Thus I should create a
blacklist of unwanted words and remove them using the `removeWords`
transformation. However, removal of these words may further mess up the
intelligibility of the original sentences. My approach is to leave them on
for the time being, then apply the blacklist when the app is complete.

One thing I've noticed is that whereas the Twitter and blogs files do
contain a fair amount of profanity, I haven't found any in the sample from
the news file, confirming my expectation that it would contain more
"polished" language.

### Whitespace cleanup

The previous transformations generate a lot of extra whitespace in the
lines of text; applying the `stripWhitespace` transformation deals with
this problem.

```{r clean_whitespace}
news_corpus_proc <- tm_map(news_corpus_proc, stripWhitespace)
```


# Data exploration

First I'll plot a histogram of the most common words remaining in the data
after the cleaning. A common approach in text mining is to create a
*term-document matrix* from a corpus: elements in this matrix represent
the occurrence of a term (a word, or an $n$-gram) in a document of the
corpus.

```{r word_dtm}
dtm <- DocumentTermMatrix(news_corpus_proc)
```

A problem with these matrices is that they tend to get very big, and they
are extremely sparse. The matrix created with the above command has only
179786 non-zero elements out of over 300 million (less than 0.06 per
cent).

The plot below shows a histogram of the ten most frequent words in the
corpus. The most frequent word is "said", perhaps not surprisingly for a
collection of news feeds (which many times will report on what someone
said about a certain issue or event).

```{r word_histogram, message=FALSE}
dtm.matrix <- as.matrix(dtm)
wordcount <- colSums(dtm.matrix)
topten <- head(sort(wordcount, decreasing=TRUE), 10)

library(reshape2)
library(ggplot2)

dfplot <- as.data.frame(melt(topten))
dfplot$word <- dimnames(dfplot)[[1]]
dfplot$word <- factor(dfplot$word,
                      levels=dfplot$word[order(dfplot$value,
                                               decreasing=TRUE)])

fig <- ggplot(dfplot, aes(x=word, y=value)) + geom_bar(stat="identity")
fig <- fig + xlab("Word in Corpus")
fig <- fig + ylab("Count")
print(fig)
```

Now I'll plot a histogram of the most frequent two-grams (pairs of words
that appear together). To construct the term-document matrix for two-grams
I use the `NGramTokenizer()` function from the `RWeka` package, which is
passed as a control term to the `DocumentTermMatrix` function.

```{r twograms}
library(RWeka)

options(mc.cores=1)
twogramTokenizer <- function(x) {
    NGramTokenizer(x, Weka_control(min=2, max=2))
}

dtm2 <- DocumentTermMatrix(news_corpus_proc,
                           control=list(tokenize=twogramTokenizer))

# I need to remove most of the sparse elements, otherwise I cannot
# allocate memory for the matrix object
dtm2_ns <- removeSparseTerms(dtm2, 0.998)
dtm2.matrix <- as.matrix(dtm2_ns)

wordcount <- colSums(dtm2.matrix)
topten <- head(sort(wordcount, decreasing=TRUE), 10)

dfplot <- as.data.frame(melt(topten))
dfplot$word <- dimnames(dfplot)[[1]]
dfplot$word <- factor(dfplot$word,
                      levels=dfplot$word[order(dfplot$value,
                                               decreasing=TRUE)])

fig <- ggplot(dfplot, aes(x=word, y=value)) + geom_bar(stat="identity")
fig <- fig + xlab("Two-grams in Corpus")
fig <- fig + ylab("Count")
print(fig)
```

Here the most frequent two-gram is "last year", and notice that "last
week" is in seventh place. Based on these results, one could imagine a
scenario where if a user inputs "last", the model predicts the most likely
completion as "year", followed by "week" (we would like the application to
output more than one suggestion for the user to choose from, so the result
should be a ranking of the 5-10 most likely terms).

## Next steps

Up to this point, the idea for predicting text would be to generate
two-gram and three-gram matrices, obtain the frequencies of the different
combinations and then match a word (or group of words) entered by the user
with the most probable $n+1$-gram.

However, I'm a bit worried about the memory requirements of this approach
- the required matrices get very large and it's very likely that for a
decent-sized training set my available computer will get overwhelmed.
Therefore I need to explore alternative approaches.


## Appendix

### Downloading the data files

The code below creates directories to store the data, if they do not exist
already, and downloads the zip file with the source data for this project.

```{r data_download}
orig_data_dir = "source_data/"
if (!dir.exists(orig_data_dir)) { dir.create(orig_data_dir) }

orig_data = "Coursera-SwiftKey.zip"
orig_file = paste0(orig_data_dir, orig_data)
if (!file.exists(orig_file)) {
  fileUrl <- paste0(
    "http://d396qusza40orc.cloudfront.net/dsscapstone/dataset/",
    orig_data)
  download.file(fileUrl, destfile=orig_file, method="curl")
}

data_dir = "data/"
if (!(dir.exists(data_dir))) {
  dir.create(data_dir)
  unzip(orig_file, exdir=data_dir)
}
```


### Obtaining file information

The following function prints some information about the data files (size,
number of lines, maximum size of a line).

```{r fileinfo, warning=FALSE, cache=TRUE}
fileInformation <- function(filepath) {
  size <- file.info(filepath)$size/1048576

  conn <- file(filepath, "r")
  fulltext <- readLines(conn)
  nlines <- length(fulltext)
  
  maxline <- 0
  for (i in 1:nlines) {
    linelength <- nchar(fulltext[i])
    if (linelength > maxline) { maxline <- linelength }
  }
  close(conn)
  
  infotext <- paste0("File: ", filepath, ", ",
                     "size: ", sprintf("%.1f", size), " MB, ",
                     "number of lines: ", nlines, ", ",
                     "max line length: ", maxline, " characters")
  
  list(size=size, nlines=nlines, maxline=maxline, infotext=infotext)
}

blog_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.blogs.txt"))
news_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.news.txt"))
twit_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.twitter.txt"))

blog_info[[4]]
news_info[[4]]
twit_info[[4]]
```


### Creating a sample of text

This code samples (approximately) a fraction of the lines of text in a
given file, chosen at random, and saves the output to another file.

```{r save_sample, warning=FALSE}
set.seed(43522)

textFileSample <- function(infile, outfile, fraction=0.5) {
  conn <- file(infile, "r")
  fulltext <- readLines(conn)
  nlines <- length(fulltext)
  close(conn)
  
  conn <- file(outfile, "w")
  selection <- rbinom(nlines, 1, fraction)
  for (i in 1:nlines) {
    if (selection[i]==1) { cat(fulltext[i], file=conn, sep="\n") }
  }
  close(conn)
  
  paste("Saved", sum(selection), "lines to file", outfile)
}

data_dir <- "data/"
samples_dir = paste0(data_dir, "samples/")
if (!(dir.exists(samples_dir))) {
  dir.create(samples_dir)
}

fraction <- 0.01
sample_file <- paste0(samples_dir, "news_sample_report.txt")
if (!file.exists(sample_file)) {
  textFileSample(paste0(data_dir, "final/en_US/en_US.news.txt"),
                 sample_file, fraction=fraction)
}
```


## References