---
title: "DSC Report"
author: "Tomás E. Tecce"
date: "March 20, 2016"
output:
  html_document:
    theme: readable
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Lorem ipsum sit dolor amet.


## Downloading the data

The data for this project comes from HC Corpora
(http://www.corpora.heliohost.org/index.html), a collection of corpora for
various languages freely available to download. The texts in the corpora
are collected from publicly available sources by a web crawler. 

The files for this project must be downloaded from the link provided in
the course website. After extracting the downloaded zip file, we end up
with four sets of three files:

- `aa_AA.blogs.txt`: text obtained from blogs
- `aa_AA.news.txt`: text obtained from news feeds
- `aa_AA.twitter.txt`: text obtained from Twitter

where `aa_AA` denotes language and locale and is either `de_DE` (German),
`en_US` (English, US), `fi_FI` (Finnish) or `ru_RU` (Russian). **For this
initial round of exploration I'll only use the English language files.**

```{r data_download}
orig_data_dir = "source_data/"
if (!dir.exists(orig_data_dir)) { dir.create(orig_data_dir) }

orig_data = "Coursera-SwiftKey.zip"
orig_file = paste0(orig_data_dir, orig_data)
if (!file.exists(orig_file)) {
  fileUrl <- paste0(
    "http://d396qusza40orc.cloudfront.net/dsscapstone/dataset/",
    orig_data)
  download.file(fileUrl, destfile=orig_file, method="curl")
}

data_dir = "data/"
if (!(dir.exists(data_dir))) {
  dir.create(data_dir)
  unzip(orig_file, exdir=data_dir)
}
```


## Selecting a data sample

The following function prints some information about the data files.

```{r fileinfo, warning=FALSE, cache=TRUE}
fileInformation <- function(filepath) {
  size <- file.info(filepath)$size/1048576

  conn <- file(filepath, "r")
  fulltext <- readLines(conn)
  nlines <- length(fulltext)
  
  maxline <- 0
  for (i in 1:nlines) {
    linelength <- nchar(fulltext[i])
    if (linelength > maxline) { maxline <- linelength }
  }
  close(conn)
  
  infotext <- paste0("File size: ", sprintf("%.1f", size), " MB, ",
                     "number of lines: ", nlines, ", ",
                     "max line length: ", maxline, " characters")
  
  list(size=size, nlines=nlines, maxline=maxline, infotext=infotext)
}

blog_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.blogs.txt"))
news_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.news.txt"))
twit_info <- fileInformation(paste0(data_dir,"final/en_US/en_US.twitter.txt"))

blog_info[[4]]
news_info[[4]]
twit_info[[4]]
```

The data files are relatively large (between 150 and 200 MB) containing
from about 900,000 lines of text in the blogs file to over 2 million in
the Twitter file. In order to speed up data exploration and the
development of an initial, prototype model, I'll use smaller samples from
each of the files.

The code below collects samples of lines at random from each file and
saves them to a new set of files, to ensure reproducibility. I select
about 20 per cent of the lines, at random, by use of the `rbinom`
function.

```{r save_sample, warning=FALSE}
set.seed(43522)

textFileSample <- function(infile, outfile, fraction=0.5) {
  conn <- file(infile, "r")
  fulltext <- readLines(conn)
  nlines <- length(fulltext)
  close(conn)
  
  conn <- file(outfile, "w")
  selection <- rbinom(nlines, 1, fraction)
  for (i in 1:nlines) {
    if (selection[i]==1) { cat(fulltext[i], file=conn, sep="\n") }
  }
  close(conn)
  
  paste("Saved", sum(selection), "lines to file", outfile)
}

samples_dir = paste0(data_dir, "samples/")
if (!(dir.exists(samples_dir))) {
  dir.create(samples_dir)
}

fraction <- 0.2
sample_file <- paste0(samples_dir, "twitter_sample.txt")
if (!file.exists(sample_file)) {
  textFileSample(paste0(data_dir, "final/en_US/en_US.twitter.txt"),
                 sample_file, fraction=fraction)
}

sample_file <- paste0(samples_dir, "news_sample.txt")
if (!file.exists(sample_file)) {
  textFileSample(paste0(data_dir, "final/en_US/en_US.news.txt"),
                 sample_file, fraction=fraction)
}
  
sample_file <- paste0(samples_dir, "blogs_sample.txt")
if (!file.exists(sample_file)) {
  textFileSample(paste0(data_dir, "final/en_US/en_US.blogs.txt"),
                 sample_file, fraction=fraction)
}
```


## Data preprocessing

```{r}
library(tm)

createCorpus <- function(filepath) {
  conn <- file(filepath, "r")
  fulltext <- readLines(conn)
  close(conn)
  
  vs <- VectorSource(fulltext)
  
  Corpus(vs, readerControl=list(readPlain, language="en", load=TRUE))
}

twit_corpuse
```

